Verification of Stencil Computations through Spatial Specifications

### 2016-07-06 paper.tex
completed soundness subsection

\begin{definition}[Schematic]
The partial function $\textsf{schematic}$ maps 
source-language indexing terms $(e_1, \ldots, e_n)$
into $n$-dimensional index schemes 
$\in (\domainVal)^n$, defined:
%
\begin{align*}
\mathsf{schematic}(e_1, \ldots, e_n) & = 
[\downarrow\!e_1 \ldots \downarrow\!e_n ] 
\quad \textit{iff} \; \forall i . \downarrow\!e_i \neq \bot 
\end{align*}
%
The $\downarrow$ operator is a partial\footnote{We represent
  partiality explicitly in our types with the set $A + \bot$.} function 
 $\downarrow : \textit{expr} \rightarrow (\domainVal + \bot)$
which maps each individal indexing expression to a component of
the indexing scheme, defined: 
%%
\begin{align*}
\downarrow\!e
 & =  \begin{cases}
a & e \equiv i + a \; \wedge \; \mathsf{neigh}(i + a) \\
-a & e \equiv i - a \; \wedge \; \mathsf{neigh}(i - a) \\
\infty & \textit{$\mathsf{IV}(e) = \emptyset$} \\
\bot   & \textit{otherwise}
\end{cases}
\end{align*}
%%
where $\mathsf{IV}(e)$ is the set of free induction variables
appearing in an expression, and $\neigh{}$ classifies expressions which 
are constant offsets from an induction variable
(Defn.~\ref{def:neighbour}). In the third case, 
if the expression is not a neighbourhood index but is not
a relative index either (\ie{} is not defined in terms
of an induction variable) then it is mapped to $\infty$, otherwise
$\downarrow\!e$ is undefined. The $\mathsf{schematic}$ function
is undefined if $\downarrow$ of any of its components $e_i$ is
undefined.

For example $\textsf{schematic}(i, j+1) = \vtwoh{0}{1}$ assuming
$i$ and $j$ are induction variables and $\textsf{schematic}(i-1, 2, j)
 = \vthreeh{-1}{\infty}{0}$. However, $\textsf{schematic}(i*2, j)$, for
 example, is
 undefined as its first index is a non-neighbour relative index, which 
 is outside the scope of our specification language.
\end{definition}

@@

\begin{example}
Take three array indexing terms
$(i, j+1)$, $(1, j+1)$ and $(3, j)$.
As given by \textsf{schematic}, these have the 
corresponding schemes: $\vtwoh{0}{1}$, $\vtwoh{\infty}{1}$
and $\vtwoh{\infty}{0}$ respectively. The following
relationship of generalises then holds:
%\[
%\xymatrix@R=0.2em@C=2em{
%[\infty \;\; 0] \ar@{-}[dr]^{\preceq_n} &  \\
%& [0 \;\; 1]  \\
%[\infty \;\; 1]  \ar@{-}[ur]_{\preceq_n} & 
%}
%\]
%%
\begin{equation*}
\vtwoh{\infty}{0} \; \not\preceq_2 \; \vtwoh{0}{1} 
\qquad
\vtwoh{\infty}{1} \; \preceq_2 \; \vtwoh{0}{1} 
\end{equation*}
%%
\end{example}
\noindent
Our terminology of \emph{indexing schemes} and
\emph{generalisation} are by analogy
with the \emph{type schemes} and \emph{type generalisation} of
polymorphic types in ML~\cite{milner1978theory}. Our notion is
much more rigid however, as our schemes do not bind names
and thus generalisation of does not require substitution; $\infty$ acts
as a positional placeholder.

@@

\begin{example}
For a $2$-dimensional situation, 
\begin{align*}
& \interp{\stenCenS{2}{1}}_2 \\
= \; & \{i\singleEntry{1}{2} \,|\, i \in \{\text{-}2,..., 2\}\!\setminus\!0 \}
  \cup  \{ \zeroEntry{1}{2} \,|\, \textit{true} \} \\
= \; & \{-2\singleEntry{1}{2}, \, -\singleEntry{1}{2}, \,
  \singleEntry{1}{2}, \, 2\singleEntry{1}{2}\} \cup
  \{\zeroEntry{1}{2}\} \\
= \; & \{\vtwoh{-2}{\infty}, \vtwoh{-1}{\infty}, \vtwoh{0}{\infty}, 
\vtwoh{1}{\infty}, \vtwoh{2}{\infty}\} 
\end{align*}
Note the absboring behaviour of $\infty$ with respect to
multiplication. 
\end{example}
\noindent
The model of $\term{+}$ is the straightforward union of 
the models of subterms. The model of $\term{*}$ is
the tensor $\otimes$ of subterm models. The tensor 
is non-trivial; we give some informal intutition first.

@@

\begin{theorem}[Soundness of approximations]
\[
\overline{\texttt{v} : S} <: \overline{\texttt{u} : T}
\; \Rightarrow \;
\interp{\overline{\texttt{v} : S}} \subseteq \interp{\overline{\texttt{u} : T}}
\]
\end{theorem}

>>>>

\begin{definition}[Schematic]
The partial function $\textsf{schematic}$ maps 
source-language indexing terms $(e_1, \ldots, e_n)$
into $n$-dimensional index schemes 
$\in (\domainVal)^n$, defined:
%
\begin{align*}
\mathsf{schematic}(e_1, \ldots, e_n) & = 
[\downarrow\!e_1 \ldots \downarrow\!e_n ] 
\quad \textit{iff} \; \forall i . \downarrow\!e_i \neq \bot 
\end{align*}
%
The $\downarrow$ operator is a partial\footnote{We represent
  partiality explicitly in our types with the set $A + \bot$.} function 
 $\downarrow : \textit{expr} \rightarrow (\domainVal + \bot)$
which maps each individal indexing expression to a component of
the indexing scheme, defined: 
%%
\begin{align*}
\downarrow\!e
 & =  \begin{cases}
a & e \equiv i + a \; \wedge \; \mathsf{neigh}(i + a) \\
-a & e \equiv i - a \; \wedge \; \mathsf{neigh}(i - a) \\
\infty & \textit{$\mathsf{IV}(e) = \emptyset$} \\
\bot   & \textit{otherwise}
\end{cases}
\end{align*}
%%
where $\mathsf{IV}(e)$ is the set of free induction variables
 in an expression and $\neigh{}$ classifies expressions which 
are constant offsets on an induction variable
(Defn.~\ref{def:neighbour}). In the third case, 
if the expression is not a neighbourhood index but is not
a relative index either (\ie{} is not defined in terms
of an induction variable) then it is mapped to $\infty$, otherwise
$\downarrow\!e$ is undefined. The $\mathsf{schematic}$ function
is undefined if $\downarrow$ of any of its components $e_i$ is
undefined.

For example, assuming
$i$ and $j$ are induction variables and $\textsf{schematic}(i-1, 2, j)
 = \vthreeh{-1}{\infty}{0}$. However, $\textsf{schematic}(i*2, j)$, for
 example, is
 undefined as its first index is a non-neighbour relative index, which 
 is outside the scope of our specification language.
\end{definition}

@@

\begin{example}
Three array indexing terms
$(i, j+1)$, $(1, j+1)$ and $(3, j)$
are mapped by \textsf{schematic} to the 
index schemes: $\vtwoh{0}{1}$, $\vtwoh{\infty}{1}$
and $\vtwoh{\infty}{0}$ respectively. The following
relationship of generalises then holds:
%\[
%\xymatrix@R=0.2em@C=2em{
%[\infty \;\; 0] \ar@{-}[dr]^{\preceq_n} &  \\
%& [0 \;\; 1]  \\
%[\infty \;\; 1]  \ar@{-}[ur]_{\preceq_n} & 
%}
%\]
%%
\begin{equation*}
\vtwoh{\infty}{0} \; \not\preceq_2 \; \vtwoh{0}{1} 
\qquad
\vtwoh{\infty}{1} \; \preceq_2 \; \vtwoh{0}{1} 
\end{equation*}
%%
\end{example}
\noindent
Our terminology of \emph{indexing schemes} and
\emph{generalisation} are by analogy
with the \emph{type schemes} and \emph{type generalisation} of
polymorphic types in ML~\cite{milner1978theory}. Our notion is
much more rigid however, as our schemes do not bind names
and thus generalisation of does not require substitution; $\infty$ acts
as a positional placeholder.

@@

\begin{example}
For a $2$-dimensional situation, 
\begin{align*}
\interp{\stenCenS{2}{1}}_2 = \; & \{i\singleEntry{1}{2} \,|\, i \in \{\text{-}2,..., 2\}\!\setminus\!0 \}
  \cup  \{ \zeroEntry{1}{2} \,|\, \textit{true} \} \\
= \; & \{-2\singleEntry{1}{2}, \, -\singleEntry{1}{2}, \,
  \singleEntry{1}{2}, \, 2\singleEntry{1}{2}\} \cup
  \{\zeroEntry{1}{2}\} \\
= \; & \{\vtwoh{-2}{\infty}, \vtwoh{-1}{\infty}, \vtwoh{0}{\infty}, 
\vtwoh{1}{\infty}, \vtwoh{2}{\infty}\} 
\end{align*}
Note the absboring behaviour of $\infty$ with respect to
multiplication. 
\end{example}
\noindent
The model of $\term{+}$ is the straightforward union of 
the models of subterms. The model of $\term{*}$ is
the tensor $\otimes$ of subterm models. The tensor 
is non-trivial; we give some informal intutition first.

@@

\begin{theorem}[Soundness of inequations
  (Figure~\ref{fig:inequations})]

Let $\leq$ be an ordering on
$\textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$
and $\leq'$ be an ordering on
$\textsf{Approx}(\mathcal{P}(\domainVal{}^n))$ defined by the
following caluses
\begin{align*}
(M \leq' N \Rightarrow \textit{inj} \, M \leq \textit{inj} \, N) &
\qquad 
(\textsf{once} \, M \leq' \textsf{mult} \, M) \\
%%
%%
(M \subseteq N \Rightarrow \textit{inj}' M \leq' \textit{inj}' N)
& \;\;\; (\textsf{exact} M \leq  \textsf{lower} M)
\;\;\; (\textsf{lower} M \leq  \textsf{upper} M)
\end{align*}
%%
where $\textit{inj}$ ranges over the unary injections
of $\textsf{Approx}$ and $\textit{inj}'$ ranges over the injections of
$\textsf{Mult}$. Then, soundness of $<:$ is that:
%%
\[
S <: T
\; \Rightarrow \;
\interp{S}_n \leq \interp{T}_n
\]
\end{theorem}
